Seeq: a library for inexact DNA sequence matching

### 2015-04-01 doc/document.tex
Added funding and corrected typos. Ready to send.

Illumina reads typically have a 1-2\% error rate \citep{Mar05},
mainly consisting of substitutions \citep{Doh08}, which are ideal
conditions for the Boyer-Moore algorithm. However, third generation
sequencing technologies such as the PacBio platform now come of age.
They produce reads two orders of magnitude longer than Illumina, at
the expense of a 10-fold higher error rate (15\%) consisting mostly
of insertions and deletions \citep{Eid09}.  
This high error rate dramatically increases the false
positive rate unless the pattern sequence is extended
accordingly. This raises a new scenario with long patterns and
large edit distances.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.6]{Figure_1.pdf}}
\caption{Dynamic DFA construction process for T='TGTGCA' and P='GCAC'.
a) Dynamic programming matrix of the alignment. The differential code
of each column is shown inside the semicircles. b) State of the DFA at
the end of the alignment.}\label{matrix}
\end{figure}

The DFA construction process is illustrated in Figure. \ref{matrix}. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{enumerate}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then set $S = S_q$.
\item Otherwise, revert $\mathbf{c}$ from $\mathbf{s}$ and compute
the $j$-th column of the edit matrix as $\mathbf{c}_j = f(\mathbf{c}, T_j)$ and
its diferential code $\mathbf{s}_j$. Then: a) If $\mathbf{s}_j$
represents an existing state ($S_q$), connect $S$ to it with a path
labeled $T_j$ and set $S = S_q$. b) If none of the states is
represented by $\mathbf{s}_j$, create $S_j$, connect $S$ to it with a
path labeled $T_j$ and set $S = S_j$.
\end{enumerate}

After several steps, the graph of the DFA is densely connected and
the steps of state and edge creation are almost never performed. 
The complexity of the algorithm is then reduced to direct DFA state
transitions.

@@

The running time of each execution is plotted against the relative
error tolerace ($k/m$) on figure \ref{fig:results}. For low error
rates, both agrep and nr-grep perform better than seeq, with a maximum
5-fold advantage when $k=1$ and $m > 25$. On the other hand, seeq is
up to two orders of magnitude faster than nr-grep and one order of
magnitude faster than agrep for $k/m > 0.15$. The similarity between
the two datasets indicates that neither software is favored when the
data has a repetitive structure. Also note that seeq 
achieves near-constant running time.

@@

In practice, most pattern searches in DNA reads are performed from
scripts and pipelines. To ease the use of seeq in existing software,
we have ensured that it is comliant with Linux pipes. Also, we have
created a Python module that can be imported and integrated seamlessly,
as it can be called on Python string objects that can be further
manipulated. We also provide an API in C for the cases that optimal
performance is required.

\paragraph{Funding\textcolon}

>>>>

Illumina reads typically have a 1-2\% error rate,
mainly consisting of substitutions \citep{Doh08}, which are ideal
conditions for the Boyer-Moore algorithm. However, third generation
sequencing technologies such as the PacBio platform now come of age.
They produce reads two orders of magnitude longer than Illumina, at
the expense of a 10-fold higher error rate (15\%) consisting mostly
of insertions and deletions \citep{Eid09}.  
This high error rate dramatically increases the false
positive rate unless the pattern sequence is extended
accordingly. This raises a new scenario with long patterns and
large edit distances.

@@

\begin{figure}[!tpb]
\centerline{\includegraphics[scale=0.6]{Figure_1.pdf}}
\caption{Dynamic DFA construction process. In this example, the search
text is TGTGCA and the pattern is GCAC.
A) Dynamic programming matrix of the alignment. The differential code
of each column is shown inside the semicircles. B) State of the DFA at
the end of the alignment.}\label{matrix}
\end{figure}

The DFA construction process is illustrated in Figure \ref{matrix}. The
dynamic programming algorithm starts at the DFA state
$S = S_0$. For any input character $T_j$, being $S$
the active DFA state, the algorithm recursively proceeds as follows:
\begin{enumerate}
\item If $S$ has an outgoing path (to $S_q$) with
label $T_j$, then set $S = S_q$.
\item Otherwise, revert $\mathbf{c}$ from $\mathbf{s}$ and compute
the $j$-th column of the edit matrix as $\mathbf{c}_j = f(\mathbf{c}, T_j)$ and
its diferential code $\mathbf{s}_j$. Then: a) If $\mathbf{s}_j$
represents an existing state ($S_q$), connect $S$ to it with a path
labeled $T_j$ and set $S = S_q$. b) If none of the states is
represented by $\mathbf{s}_j$, create $S_j$, connect $S$ to it with a
path labeled $T_j$ and set $S = S_j$.
\end{enumerate}

After several steps, the graph of the DFA is densely connected and
case 2 almost never occurs. The complexity of the algorithm is then
reduced to direct DFA state transitions.

@@

The running time of each execution is plotted against the relative
error tolerace ($k/m$) on Figure \ref{fig:results}. For low error
rates, both agrep and nr-grep perform better than seeq, with a maximum
5-fold advantage when $k=1$ and $m > 25$. On the other hand, seeq is
up to two orders of magnitude faster than nr-grep and one order of
magnitude faster than agrep for $k/m > 0.15$. The similarity between
the two datasets indicates that neither software is favored when the
data has a repetitive structure. Also note that seeq 
achieves near-constant running time.

@@

In practice, most pattern searches in DNA reads are performed from
scripts and pipelines. To ease the use of seeq in existing software,
we have ensured that it is compliant with Linux pipes. Also, we have
created a Python module that can be imported and integrated seamlessly,
as it can be called on Python string objects that can be further
manipulated. We also provide an API in C for the cases that optimal
performance is required.

\paragraph{Funding\textcolon}
The research leading to these results has received funding from the
European Research Council under the European Union's Seventh Framework
Programme (FP7/2007-2013) / ERC SYNERGY n$^\circ$ 609989.
